Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-09-25 paper/paper.tex
wrapfig
wip

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage{iclr2020_conference,times}
\DeclareMathOperator*{\argmax}{argmax} % thin space, limits underneath in displays

@@

     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
     %
     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
\end{theorem}

\begin{proof}
  See Appendix \ref{appendix:theorem_proof}.
\end{proof}

% \begin{proof}
%
%  Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
%  \begin{align*}
%     \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
% %                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
%                                                         &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
%      \label{eq:doublenorm}
% \end{align*}
% \begin{align*}
%     \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
%     &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
% %    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
%     &= EF \hat G
% \end{align*}
%
% Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero. Let $Z=F^\dagger EF\hat G$. We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$). Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$. Therefore,
%  $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
%
% But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$. Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$. Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$. This completes the proof.
% \end{proof}

So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.

@@

$X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right)$, $N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right)$ and

$\hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right)$

$Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1})$

In the absence of noise, we have $Diag(\hat H) = Diag(E)$. Else the k first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.

The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by k. On average, since $X$ and $N$ are not correlated by definition, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition numbers, and $\frac{Var(X)}{Var(X)+Var(N)}$ provide a good estimate of $\frac{1}{k}Tr(\hat H)$.

>>>>

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref, amsthm}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage{iclr2020_conference,times,wrapfig}

\DeclareMathOperator*{\argmax}{argmax} % thin space, limits underneath in displays

@@

     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
     %
     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimize
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
\end{theorem}
%
\begin{proof}
  See Appendix \ref{appendix:theorem_proof}.
\end{proof}
%
% \begin{proof}
%
%  Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
%  \begin{align*}
%     \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
% %                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
%                                                         &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
%      \label{eq:doublenorm}
% \end{align*}
% \begin{align*}
%     \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
%     &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
% %    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
%     &= EF \hat G
% \end{align*}
%
% Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero. Let $Z=F^\dagger EF\hat G$. We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$). Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$. Therefore,
%  $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
%
% But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$. Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$. Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$. This completes the proof.
% \end{proof}
%
\begin{equation}
  \text{So, }\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top.
\end{equation}
% So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.

@@

% $$
\begin{equation}
  X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right),
\end{equation}
% $$
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right),
\end{equation}
\begin{equation}
  Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1}).
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $Diag(\hat H) = Diag(E)$. Otherwise the $k$ first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.

The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by $k$. On average, since $X$ and $N$ are not correlated by definition, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition number, and $\frac{Var(X)}{Var(X)+Var(N)}$ provides a good estimate of $\frac{1}{k}Tr(\hat H)$.
